---
title: "FIFA2020 Analysis"
author: "Oh Jian Hui"
date: "17/10/2020"
output:
  html_document:
    toc: true
---

```{r import, include=FALSE}
library(tidyverse)
library(readxl)
```

```{r datasets, include=FALSE}
track <- read_excel("../data/player_tracking_fifa2018.xlsx") %>%
  mutate(match_id = as.character(match_id))
info <- read_excel("../data/player_info_fifa2018.xlsx")
sched <- read_excel("../data/match_schedule.xlsx") %>%
  mutate(date_time = as.POSIXct(date_time, format = "%d %b %Y - %H:%M"))
```

# Question 1

```{r qn1, echo=FALSE}
# Outfield players
outfield_list <- info %>% filter(Pos != "GK")

# Parse time_played column
seconds_col <- str_split(track$time_played, pattern = "('|\")") %>%
  {do.call(rbind, .)[, -3]} %>%
  {as.numeric(.[, 1]) * 60 + as.numeric(.[, 2])}
track2 <- mutate(track, time_played = seconds_col)

# Outfield players who played at least 30 minutes
df1 <- track2 %>%
  right_join(outfield_list, by = c("jersey" = "Jersey", "team" = "Country")) %>%
  filter(time_played >= 30 * 60)

# Median top speed for each team, for each position
df1a <- df1 %>%
  group_by(team, Pos) %>%
  summarise(med_speed = median(top_speed), .groups = "drop")

# Helper function to check whether midfielder median top speed greater than forward
super_mid <- function(pos, speed) {
  mid <- speed[pos == "Midfielder"]
  fwd <- speed[pos == "Forward"]
  if(mid > fwd) {
    return(TRUE)
  }
  return(FALSE)
}

# Data frame for plotting
df1b <- df1a %>%
  mutate(Pos = factor(Pos, levels = c("DF", "MF", "FW"),
                         labels = c("Defender", "Midfielder", "Forward"))) %>%
  group_by(team) %>%
  mutate(super_mf = super_mid(Pos, med_speed))

# Text labels
df1_labels <- df1b %>% filter(super_mf, Pos == "Forward")

# Plot
ggplot(data = df1b) +
  geom_line(aes(x = Pos, y = med_speed, group = team, colour = super_mf),
            size = 0.5, show.legend = FALSE) +
  geom_text(data = df1_labels, mapping = aes(x = Pos, y = med_speed, label = team),
            nudge_x = 0.1, size = 3) +
  labs(title = "Median Top Speed by Position",
       subtitle = "Most teams' forwards sprint faster than their midfielders, but not all..",
       x = "Position", y = "km/h")
```

```{r clear, include=FALSE}
# Clear up workspace
rm(list = setdiff(ls(), c("track2", "sched", "info")))
```

# Question 2

### Time spent by defenders on the oposition half in matches of different stages

For this question, I only consider matches played by teams that have at least reached the knock-out stage. Otherwise, it would be impossible compare whether the teams decided to switch to play more conservatively during the knock-out matches.

Also, only defenders who played at least 45 minutes are considered. This is because a previous plot I made shows that some defenders playing less than 45 minutes clustered far from the rest (but not out of the 1.5 * IQR because they reached the limit of the y-scale), extending the whiskers' length.

```{r qn2, echo=FALSE}
# Parse the column time_oh
if_else_vec <- Vectorize(if_else)

track3 <- track2 %>%
  separate(time_oh, into = c("time_oh_pc", "_")) %>%
  select(-`_`) %>%
  mutate(time_oh_pc = as.numeric(time_oh_pc),
         time_oh_pc = if_else_vec(is.na(time_oh_pc), 0, time_oh_pc))
# to my understanding, missing value implies 0 time in the opposition half

# Teams who reached knock-out stage
ko_teams <- sched %>%
  filter(match_info %in% c("Round of 16", "Quarter-finals", "Semi-finals",
                           "Play-off for third place", "Final")) %>%
  {union(unique(.$home_team), unique(.$away_team))}

track4 <- track3 %>% filter(team %in% ko_teams)

# Identify match id of knock-out stage and group stage
ko_matches <- sched %>%
  filter(match_info %in% c("Round of 16", "Quarter-finals", "Semi-finals",
                           "Play-off for third place", "Final")) %>%
  {unique(.$match_id)}

"%!in%" <- function(x, y) !("%in%" (x, y))
grp_matches <- sched %>%
  filter(match_info %!in% c("Round of 16", "Quarter-finals", "Semi-finals",
                            "Play-off for third place", "Final")) %>%
  {unique(.$match_id)}

# Defenders
defenders_list <- info %>% filter(Pos == "DF")

# Find average time of defenders on opposite half
# use the average percentage of time instead of absolute time
pc_time_opp <- function(stage) {
  track4 %>%
    right_join(defenders_list, by = c("jersey" = "Jersey", "team" = "Country")) %>%
    filter(match_id %in% stage, !is.na(time_played), time_played >= 45 * 60) %>%
    summarise(average_in_percent = mean(time_oh_pc), .groups = "drop")
}

# Average in group stage matches
print("Group Stage matches")
(time_gs <- pc_time_opp(grp_matches))

# Average in knock-out matches
print("Knock-out matches")
(time_ko <- pc_time_opp(ko_matches))
```

```{r qn2plot, echo=FALSE}
# Plotting
set.seed(2020)
to_plot <- track4 %>%
  right_join(defenders_list, by = c("jersey" = "Jersey", "team" = "Country")) %>%
  filter(!is.na(time_played)) %>%
  mutate(stage = if_else_vec(match_id %in% ko_matches, "Knock-Out Stage", "Group Stage")) %>%
  filter(time_played >= 45 * 60)

# Graphics
to_plot %>%
  ggplot(aes(x = stage, y = time_oh_pc)) +
  geom_boxplot(outlier.alpha = 0, colour = "steelblue") +
  geom_jitter(height = 0, width = 0.1, alpha = 0.5) +
  # Average lines and labels
  geom_segment(aes(x = 0.6, y = unlist(time_gs), xend = 1.4,
                             yend = unlist(time_gs)), colour = "green4", size = 1) +
  geom_segment(aes(x = 1.6, y = unlist(time_ko), xend = 2.4,
                   yend = unlist(time_ko)), colour = "green4", size = 1) +
  annotate("text", x = 1.48, y = unlist(time_gs) + 1,
           label = str_c("Average\n", "28.9"), colour = "green4", size = 3) +
  annotate("text", x = 2.48, y = unlist(time_ko) + 1,
           label = str_c("Average\n", "26.5"), colour = "green4", size = 3) +
  # Title
  labs(title = "Time that Defenders Spend on the Opposition Half",
       subtitle = "in each match in different stages",
       x = NULL, y = "%",
       caption = "*only consider teams that reached the knock-out stage
       and defenders who played at least 45 minutes")
```

Indeed, in general, the teams switched to play more conservatively during the knock-out stage.  
Since mean is higher than median in general, the two graphs are right-skewed. In group stage matches, the skewness is much higher and the spread is also perceptibly higher.

```{r qn3plot2, echo=FALSE}
to_plot2 <- to_plot %>%
  group_by(team, stage) %>%
  summarise(med_time = median(time_oh_pc), .groups = "drop")

conser <- to_plot2 %>%
  pivot_wider(names_from = "stage", values_from = "med_time") %>%
  mutate(cons = if_else_vec(`Group Stage` > `Knock-Out Stage`, "Y", "N"))

lab1 <- conser %>%
  filter(cons == "N") %>%
  mutate(team = recode(team, BEL = "BEL\nDEN", DEN = ""))
lab2 <- conser %>%  filter(cons == "Y")

to_plot2 %>%
  left_join(conser, by = "team") %>%
  ggplot() +
  geom_line(aes(x = stage, y = med_time, group = team, colour = cons),
            show.legend = FALSE) +
  geom_text(data = lab1, aes(x = 2.05, y = `Knock-Out Stage`, label = team),
            size = 2.5) +
  geom_text(data = lab2, aes(x = 0.95, y = `Group Stage`, label = team),
            size = 2.5) +
  labs(title = "Difference of Defenders' Time on Opposite Half",
       subtitle = "by Country and Stage",
       x = "", y = "Median Time (in %)")
```

Out of the 16 teams considered, there are 7 teams (in red) that play less conservatively during the high stake Knock-Out Stage. POR, SUI and CRO in particular, have especially high difference in playing style (noted by the high slope)